Physics-based protein structure prediction and design using the confinement method

### 2012-10-29 Paper/paper.tex
Some edits on text

The calculation of free energy differences is of central importance in the simulation of biochemical
systems.  The computation of the free energy difference between pairs of macromolecule with large
conformational change is particularly a difficult task and computationally expensive with existing
methods. In this work, an improved version of the confinement approach is used to calculate absolute
free energies of bio molecular systems. The method does not require a reaction coordinate or
transition path. It is fast to compute. We show that the method correctly picks out the state of
lowest free energy of a pair of structures having similar sequence but different fold. We also show
using models from CASP9 that the method picks out native-like structures from misfolded or decoys,
provided at least one good structure is in the input set.

@@

In general, proteins with similar sequences tend to have similar structures. This idea is the basis
of comparative modeling and fold recognition in protein structure prediction. There are, however,
examples---often referred to as chameleon sequences---of proteins with similar sequences that have
remarkably different structures. Orban and co-workers have designed a series of 56-residue proteins
(based on Protein G) that adopt one of two different folds depending on small changes in sequence
(see Figure~\ref{fig:orban}). Sequences that adopt a mixed alpha/beta structure similar to Protein G
are denoted as ``GB'', while sequences that form a three-helix bundle are denoted as ``GA''. One
pair of sequences (GA88/GB88) are 88 percent identical and differ in seven positions. Another pair
(GA95/GA95) are 95 percent identical and differ in three positions. The last pair (GA98/GB98) differ
only in a single tyrosine to alanine mutation.

@@

We initially approached this problem by making a model of each sequence with the same backbone
structure as its partner chameleon sequence. For example, we took the sequence of GA88 and built a
model with the same overall structure as GB88. We then used the confinement method to assess the
free energy difference between the experimentally determined structure of GA88 and the model (with
the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all six sequences (data not shown). There is, however, a
serious problem with this analysis: we are comparing an experimentally determined structure with a
computer generated model. It might be possible that we were able to make correct predictions simply
because artefacts of our modeling procedure always lead to the model having a higher free energy
than the experimental structure.

@@

The results of these calculations are presented in Figure~\ref{fig:orban}. The confinement method
identifies the correct structure for all five sequences. One of the hypothesis for fold switching
of chamelon sequence is that the structural transitions require states with diminished stability.
It is widely believed that if the free energy of the native state and the alternative state is within
 a range of $-5 Kcal/Mol$ then  it can quickly change fold when the stability of the native state 
decreases. The stability of the native state can decrease for a number reason ranging from chemical 
modification, breaking of disulphide bond or mutation as in this case. The free energy differences, that
came out from our calculation ranges from
around 2.9 to 5.7 kcal/mol, which is consistent with the above hypothesis \cite{He2008}, \cite{{Alexander2009}}.

@@

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test world wide competition in which different groups apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. Our second test was to participate as a
metapredictor group in CASP10. In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}. 
GDT\_TS represent the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).

@@

To further test the method we have taken the example of protein BT2368 from Bacteroides thetaiotaomicron. The PDB id of this 
74 amino acid residue protein is 2L02 and CASP target code is T0560. We have compared the free energy difference between 
the native structure and the two of the five submitted models from the group "Splicer". The remaining three models are discarded as 
again they are similar to the rest of the models. The comparison of GDT\_TS score as shown in 
Figure~\ref{fig:T0560} and published in final CASP result is only for models with residues 3 to 66. In order to keep
consistency, we have also do our analysis with models and crystal structure consisting residues 3 to 66. As expected,
the native state is identified correctly, and the two other models is ranked in the correct order.

@@

There are however some instances where the method can fail, specially when GDT scores are
very close. We have found this for an engineered protein from Asr4154 protein (PDB ID: 2L09 and CASP Target T0538).
This protein contains 54 amino acid residues with three alpha helixes and two very short beta strand. 
The model that is closest to the native structure was generated by the group PconsR with a GDT\_TS value of 96.23.
We have also considered one model from group "Shell" (GDT\_TS = 90.09) and "FOLDIT" (GDT\_TS = 86.32) for our analysis along with
the native crystal struture. The models from PconsR, Shell and FOLDIT are labelled as model 1, model 2 and model 3 
respectively in Figure~\ref{fig:T0538}. As presented from Figure~\ref{fig:T0538}, one can found that the 
model from PconsR has lowest free energy value, making it a more stable structure. Although ranking of the other models
remain consistent, it may happen that the theoretical model produced by the group "PconsR" much more stable than the 
crystal structure. This is particularly difficult target for rank ordering as all the models those are considered for this 
calculation are within RMSD values of 2.6 \AA.

@@

We mentioned earlier that our predictive abilities are greatly decreased with low model qualities. It is
better to have atleast one good model to pick out the native/native like structures from the rest of the decoy. 
In oorder to test this hypothesis we performed another test with target T0531 of CASP9, in which we compared the crystal 
structure to five models from the MUFOLD server. The pdb id of this extracellular domain of the jumping translocation 
breakpoint protein is 2KJX and the CASP target code is T0531.
Looking at Figure~\ref{fig:T0531} shows: 1.) native is correctly
identified as expected and 2.) Surprisingly there is a high level of correlation between the GDT
ranking and the free energy ranking, only two structures with GDT\_TS score less than  35  are ordered
incorrectly. It is worth to note that models 2 and 3 have the same GDT and very different free
energies, meaning that the actual ordering could change a lot \cite{Perez2012}.

@@

It is often difficult to model CASP targets using confinement and release method. There are many cases,
where the sequence of a particular target are trimmed during actual result presentation. This is illustrated
with the crystal structure of the leucine zipper domain of cGMP dependent protein kinase I (pdb id: 3NMD
and CASP9 code: T0605) as shown in Figure~\ref{fig:T0605}. The actual CASP 9 target consisted of 72 amino acid residues, but
as only 18-66 amino acid residues could be detected in actual crystal structure of this protein, the final result was trimmed 
with respect to that crystal structure with 49 amino acid residues which consist of a single alpha helix. The amino acid residues 18 and 66 are shown in
three models submitted by the group "Baker". It is important to note that blind prediction of model ranking on the basis of free energy can be hugely 
affected in such scenario as the only major difference between those models are the trimmed parts, specially the region consisting residues 1-18.

\subsubsection{Can confinement method perform better quality assessment in protein structure prediction ?}
A part of CASP experiment is dedicated to the assessment of the quality of predicted model \cite{Kryshtafovych2011}. 
Most of the top performing group in this catagory use consensus approaches for quality assessment \cite{Wang2011}.
A review of CASP9 quality assessment also pointed out that the method which are based on clustering technique perform
much better compare to those methos which are based on analysis of individual model \cite{Kryshtafovych2011}. 
In this background we try to answer, whether confinement method can do better quality assessment compare to the best performing
group (MUFOLD-WQA) in this catagory.
We have again picked up the CASP taget T0538 to answer this question. We have chosen the Model 3 submitted by PconsR (GDT\_TS = 96) 
and model 5 from the MULTICOM-NOVEL ( GDT\_TS = 83). Both are server predicted model and they were further used for quality assessment 
by the group MUFOLD-WQA \cite{Wang2011}. It is important to note that the model from PconsR was most accurate predicted model for this 
target and the model from MULTICOM-NOVEL was predicted as best by the quality assessment experiments of MUFOLD-WQA. 
Our free energy analysis we have shown that model from Pconsr is better than the model from 
MULTICOM-NOVEL $(\Delta \Delta G (Model\_Pconsr - Model\_MULTICOM-NOVEL = -8.2 Kcal/Mol)$, which   
matches well with the GDT\_TS score. On the other hand the MUFOLD-WQA predicted the ranking in the reverse order. 
There is no doubt that the consensus approaches can predict the model quality in a faster manner. But we expect that confinement
approach can predict the model quality in a relatively expensive but much accurate way.

@@

\subsubsection{CASP10}

We participated in CASP10 as a metapredictor group. Given the tigh schedule of the CASP competition,
we only attempted proteins under 100 residues in length. Our main problem during this part of the
competition. was that proteins were trimmed from the actual calculation.
%\Alberto{Justin,Arijit: Probably should not comment on metapredicting since we were not able to do
%    good due to the trimming factor. Instead, we should move forward with Justin's idea of using
%    structures that other meta predictors did badly at. What stage is that in?}

@@

Proteins are essential parts of organisms and participate in virtually every process within living cells. 
In order to understand different functions of proteins, detect diseases caused by them and design more efficient drugs 
it is often useful to have the three dimensional structure of the protein. Experimental studies using NMR and X-ray crystallography is 
not enough to detect large number of structures. In recent past there are number of studies directed towards understanding protein 
structure using theoretical tools. Experimentally, it is relatively easy to identify the native structure from the data obtained 
from x-ray crystallography or NMR. On the other hand using theoretical tools it is often difficult to identify the native structure 
using both physics based methods and knowledge based method. In order to rank structure from the available data, some groups use 
clustering algorithms and the structure of the
most populated cluster is picked out as the most probable structure. However, it may happen that the structure may be trapped and 
spend most of its time in a kinetic trap, which may give a wrong interpretation of the clustering result. On the other hand, some groups
also depend on potential energy function to pick out their best structure. In this paper, we have used a free energy method to identify
the native/native like structure. First, we have tested this method for a chameleon sequence, which has a $95$ percent sequence identity
but one of them has a $3 \alpha$ structure and another one has $4 \beta + \alpha$ structure. Using the current method, we can found that 
the native structure always has most stable structure. We have also tested this method for different targets of CASP experiments. We found
that in most cases, it can pick out the native/native like structure in terms of lowest free enrgy value.

However, there may be some limitation of this method in terms of size of the protein molecule. The normal mode analysis may not be
accurate for proteins larger than 100 amino acid residues. This can be approximated by removing the hydrogen molecules from 
the system. 
The method can be also give wrong result if all the models in the decoy are very poor. It may also be difficult to rank structures if 
all the structures are very very close to the native structure. But that does not matter at all as our main aim is to differentiate
between the good and the bad structure. In this article we have extensively tested this method using examples from CASP experiment.
This method can be computationally expensive in actual CASP experiment if one want rank thousands of submitted models in CASP 
experiments. However, in real life experiment with a particular target this method can give accurate result. The important 
fact about Confinement method is that all the part of the calculation can be done independently. We have mostly carried out 
our calculation with Amber program \cite{Case2005} in GPU computer. With a avarage 56 residue we found that it take only 4 hour to complete 
20 ns of the confinement step. So, with available computer power this method will be easy to use.

@@

>>>>

The calculation of free energy differences is of central importance in the simulation of biochemical
systems.  The computation of the free energy difference between pairs of macromolecular
conformations is currently a particularly difficult and computationally expensive task with existing
methods. In this work, an improved \Alberto{what is the improvement} version of the confinement approach is used to calculate absolute
free energies of bio molecular systems. The method provides two main advantages: it does not require a reaction coordinate or
transition path and it is fast to compute. Throught out the article we show its use in different
challenging modeling problems. In particular, we show its use in predicting the conformational
preference of chamaleon sequences (sequences with high sequence identity and different
folds). This sequence dependent conformational preference set the stage for the use of the method in
protein design.

@@

In general, proteins with similar sequences tend to have similar structures. This idea is the basis
of comparative modeling and fold recognition in protein structure prediction. There are, however,
examples---often referred to as chameleon sequences---of proteins with similar sequences that have
remarkably different structures. Orban and co-workers have designed a series of 56-residue proteins
(based on Protein G) that adopt one of two different folds depending on small changes in sequence
(see Figure~\ref{fig:orban}). Sequences that adopt a mixed alpha/beta structure similar to Protein G
are denoted as ``GB'', while sequences that form a three-helix bundle are denoted as ``GA''. One
pair of sequences (GA88/GB88) are 88 percent identical in sequence and differ in seven positions. Another pair
(GA95/GA95) are 95 percent identical and differ in three positions. The last pair (GA98/GB98) differ
only in a single tyrosine to alanine mutation.

@@

We initially approached this problem by making a model of each sequence with the same backbone
structure as its partner chameleon sequence. For example, we took the sequence of GA88 and built a
model with the same overall structure as GB88. We then used the confinement method to assess the
free energy difference between the experimentally determined structure of GA88 and the model (with
the GA88 sequence and the GB88 structure). The confinement method was able to predict the
conformational preferences correctly for all six sequences (data not shown). This is however not
surprising. It is well known [CITE] that it is easy to distinguish computational models from native
structures. Despite the fact that this models are built by having a huge amount of structural
information, it might be possible that we were able to make correct predictions simply
because artefacts of our modeling procedure always lead to the model having a higher free energy
than the experimental structure.

@@

The results of these calculations are presented in Figure~\ref{fig:orban}. The confinement method
identifies the correct structure for all five sequences. One of the hypothesis for fold switching
of chamelon sequence is that the structural transitions require states with diminished stability.
It is widely believed that if the free energy of the native state and the alternative state is within
 a range of $-5 Kcal/Mol$ then  it can quickly change fold when the stability of the native state 
decreases. The stability of the native state can decrease for a number reason ranging from chemical 
modification, breaking of disulphide bonds or mutations as in this case. The free energy differences, that
came out from our calculation ranges from
around 2.9 to 5.7 kcal/mol, which is consistent with the above hypothesis \cite{He2008}, \cite{{Alexander2009}}.

@@

This method is particularly well suited to pick out the native/native like structure from misfolded or decoys. We have
tested this using different models from CASP (critical assessment of structure
prediction). CASP is a blind test world wide competition in which different groups apply their
methods to predict the 3D structure of proteins given their sequence. This is done with a strict 3
week limit on each target (3 day for servers) and each group is allowed to submit 5 possible
structures (ranked from better to worst). In our role as assessors during the CASP8 and CASP9 \cite{MacCallum2011}
competition we observed no correlation at all between the ordering of structures submited by the
groups and the real ranking compared to an experimental model \cite{Kryshtafovych2011}}. The consequences of this go beyond
those five structures; the deeper meaning is that groups producing ensembles of structures are
generating structures that are better than the ones they submit, but they do not know about it. In
fact, when querrying different groups after the results were known, most groups agree that this is
the case. Beyond the CASP problem, this reflects on the modelers ability to correctly rank order
models in many different environments, from structures for drug design leads to designing more
stable proteins or peptide mimetics such as peptoids. One of the main culprits of this lack of
accuracy is the fact that ranking is often done via a potential energy function, which in many cases
lacks an entropy component. Other initiatives including knowledge based potentials do use some sort
of free energy to rank order, but its accuracy is not enough. Our method provides a physics based
solution to this problem. We have tried two experiments centered on CASP. First we tried to rank
order some structures from the previous CASP9 experiment. Our second test was to participate as a
metapredictor group in CASP10. In both methods, our ranking is determined as a free energy measure,
and it is compared with the ranking given by a geometrical comparison between native and submitted
models called GDT\_TS (Global distance test score \cite{Zemla2003}). 
GDT\_TS represents the average percentage of residues that are in close proximity in two
structures optimally superimposed using four different distance cutoffs (1, 2, 4 and 8 Å).

@@

To further test the method we have taken the example of protein BT2368 from Bacteroides thetaiotaomicron. The PDB id of this 
74 amino acid residue protein is 2L02 and CASP target code is T0560. We have compared the free energy difference between 
the native structure and the two of the five submitted models from the group "Splicer". The remaining three models are discarded as 
again they are similar to the rest of the models. The comparison of GDT\_TS score as shown in 
Figure~\ref{fig:T0560} and published in final CASP result is only for models with residues 3 to 66. In order to keep
consistency, we have also done our analysis with models and crystal structure consisting residues 3 to 66. As expected,
the native state is identified correctly, and the two other models are ranked in the correct order.

@@

There are however some instances where the method can fail, specially when GDT scores are
very close. We have found this for an engineered protein from Asr4154 protein (PDB ID: 2L09 and CASP Target T0538).
This protein contains 54 amino acid residues with three alpha helixes and two very short beta strand. 
The model that is closest to the native structure was generated by the group PconsR with a GDT\_TS value of 96.23.
We have also considered one model from group "Shell" (GDT\_TS = 90.09) and "FOLDIT" (GDT\_TS = 86.32) for our analysis along with
the native crystal struture. The models from PconsR, Shell and FOLDIT are labelled as model 1, model 2 and model 3 
respectively in Figure~\ref{fig:T0538}. As seen in Figure~\ref{fig:T0538}, one can find that the 
model from PconsR has lowest free energy value, making it a more stable structure. Although ranking of the other models
remain consistent, it may happen that the theoretical model produced by the group "PconsR" much more stable than the 
crystal structure. This is particularly difficult target for rank ordering as all the models those are considered for this 
calculation are within RMSD values of 2.6 \AA.

@@

We mentioned earlier that our predictive abilities are greatly decreased with low model qualities. It is
better to have at least one good model to pick out the native/native like structures from the rest of the decoy. 
In order to test this hypothesis we performed another test with target T0531 of CASP9, in which we compared the crystal 
structure to five models from the MUFOLD server. The pdb id of this extracellular domain of the jumping translocation 
breakpoint protein is 2KJX and the CASP target code is T0531.
Looking at Figure~\ref{fig:T0531} shows: 1.) native is correctly
identified as expected and 2.) Surprisingly there is a high level of correlation between the GDT
ranking and the free energy ranking, only two structures with GDT\_TS score less than  35  are ordered
incorrectly. It is worth to note that models 2 and 3 have the same GDT and very different free
energies, meaning that the actual ordering could change a lot \cite{Perez2012}.

@@

It is often difficult to model CASP targets using the confinement and release method. There are many cases,
where the sequence of a particular target is trimmed during actual result presentation. This is illustrated
with the crystal structure of the leucine zipper domain of cGMP dependent protein kinase I (pdb id: 3NMD
and CASP9 code: T0605) as shown in Figure~\ref{fig:T0605}. The actual CASP 9 target consisted of 72 amino acid residues, but
as only 18-66 amino acid residues could be detected in actual crystal structure of this protein, the final result was trimmed 
with respect to that crystal structure with 49 amino acid residues which consist of a single alpha helix. The amino acid residues 18 and 66 are shown in
three models submitted by the group "Baker". It is important to note that blind prediction of model ranking on the basis of free energy can be hugely 
affected in such scenario as the only major difference between those models are the trimmed parts, specially the region consisting residues 1-18.

\subsubsection{Can confinement method perform better quality assessment in protein structure prediction ?}
A part of CASP experiment is dedicated to the assessment of the quality of predicted model \cite{Kryshtafovych2011}. 
Most of the top performing group in this catagory use consensus approaches for quality assessment \cite{Wang2011}.
A review of CASP9 quality assessment also pointed out that the methods which are based on clustering
techniques perform
much better compared to those methos which are based on analysis of individual models \cite{Kryshtafovych2011}. 
In this background we try to answer, whether the confinement method can do better quality assessment compare to the best performing
group (MUFOLD-WQA) in this catagory.
We have again picked up the CASP taget T0538 to answer this question. We have chosen the Model 3 submitted by PconsR (GDT\_TS = 96) 
and model 5 from the MULTICOM-NOVEL ( GDT\_TS = 83). Both are server predicted models and they were further used for quality assessment 
by the group MUFOLD-WQA \cite{Wang2011}. It is important to note that the model from PconsR was the most accurate predicted model for this 
target and the model from MULTICOM-NOVEL was predicted as best by the quality assessment experiments of MUFOLD-WQA. 
Our free energy analysis we have shown that model from Pconsr is better than the model from 
MULTICOM-NOVEL $(\Delta \Delta G (Model\_Pconsr - Model\_MULTICOM-NOVEL = -8.2 Kcal/Mol)$, which   
matches well with the GDT\_TS score. On the other hand the MUFOLD-WQA predicted the ranking in the reverse order. 
There is no doubt that the consensus approaches can predict the model quality in a faster manner. But we expect that confinement
approach can predict the model quality in a relatively expensive but much accurate way.

@@

@@

Proteins are essential parts of organisms and participate in virtually every process within living cells. 
In order to understand different functions of proteins, detect diseases caused by them and design more efficient drugs 
it is necessary to have the three dimensional structure of the protein. Experimental studies using NMR and X-ray crystallography is 
not able to keep up with the sheer number of protein sequences available. In recent past there are number of studies directed towards understanding protein 
structure using theoretical tools. Experimentally, it is relatively easy to identify the native structure from the data obtained 
from x-ray crystallography or NMR. On the other hand using theoretical tools it is often difficult to identify the native structure 
using both physics based methods and knowledge based method. In order to rank structure from the available data, some groups use 
clustering algorithms and the structure of the
most populated cluster is picked out as the most probable structure. However, it may happen that the structure may be trapped and 
spend most of its time in a kinetic trap, which may give a wrong interpretation of the clustering result. On the other hand, some groups
also depend on potential energy functions to pick out their best structure, lacking an entropic
component. In this paper, we have used a free energy method to identify
the native/native like structure. First, we have tested this method for a chameleon sequence, which has a $95$ percent sequence identity
but one of them has a $3 \alpha$ structure and another one has $4 \beta + \alpha$ structure. Using the current method, we found that 
the native structure is always the most stable \Alberto{This is not true!!}. We have also tested this method for different targets of CASP experiments. We found
that in most cases, it can pick out the native/native like structure in terms of lowest free enrgy value.

However, there may be some limitations to this method in terms of size of the protein molecule. The normal mode analysis may not be
accurate for proteins larger than 100 amino acid residues. This can be approximated by removing the hydrogen molecules from 
the system. 
The method can be also give wrong result if all the models in the decoy are very poor. It may also be difficult to rank structures if 
all the structures are very very close to the native structure. But that does not matter at all as our main aim is to differentiate
between the good and the bad structure. In this article we have extensively tested this method using examples from CASP experiment.
This method can be computationally expensive in actual CASP experiment if one want rank thousands of submitted models in CASP 
experiments. However, in real life experiment with a particular target this method can give accurate result. The important 
fact about Confinement method is that all the part of the calculation can be done independently. We have mostly carried out 
our calculation with Amber program \cite{Case2005} in GPU computer. With a avarage 56 residue we found that it take only 4 hour to complete 
20 ns of the confinement step. So, with available computer power this method will be easy to use.

@@

\Alberto{Somewhere in the introduction or methods}
There are two big bottlenecks in this calculation. The first has to do with the amount of
simulations that have to be performed to confine the ensemble of conformations close to a microstate
into a single microstate, the second bottleneck has to do with the size of the system in the normal
mode calculation. We have tackled the first issue by using GPU (Graphical Processing Unit) technology as opposed to the classical
CPUs, which gives us a two order of magnitude boost in computational efficiency. Normal mode
analysis calculations are Oh(N**3) and require large amounts of memory, effectively limiting the
size of the systems under study. Where as this can be overcome by larger memory machines, these are
often not available, and so we propose an alternative by using quasiharmonic analysis. This is still
an Oh(N**3) problem, but we now the motions of atoms are already determined in trajectories, and so
we can choose to do quasiharmonic analysis excluding hydrogens, which will reduce the number of
atoms N to almost N/2. This means that we can use this method for systems twice as large.
